An introduction to Docker for reproducible research, with examples from the R environment

### 2014-09-19 docker-paper/manuscript.Rmd
tweaking

Keywords:
  - docker
  - virtualization
  - reproducibility

@@

Reproducible research has received an increasing level of attention
throughout the scientific community [e.g. @Peng_2011; @Nature_2012]
and the public at large [e.g. @The_Economist_2013]. All steps of the
scientific process, from data collection and processing, to analyses,
visualizations and conclusions depend ever more on computation and
algorithms, _computational reproducibility_ has received particular
attention [@Merali_2010].  Though in principle this algorithmic dependence
should make such research easier to reproduce -- computer codes being
both more portable and potentially more precise to exchange and run
than experimental methods -- in practice this has led to an ever larger
and more complex black box that stands between what was actually done
and what is described in the literature.  Crucial scientific processes
such as replicating the results, extending the approach or testing the
conclusions in other contexts, or even merely installing the software
used by the original researchers can become immensely time-consuming if
not impossible.

@@

Systems research has long concerned itself with the issues of
computational reproducibility and the technologies that can facilitate
those objectives [e.g. @Eide_2010, @Harji_2013]. Docker is a new but already
very popular open source tool that combines many of these approaches
in a user friendly implementation, including: (1) performing Linux
container (LXC) based operating system (OS) level virtualization, (2)
portable deployment of containers across platforms, (3) component reuse,
(4) sharing, (5) archiving, and (6) versioning of container images. While
Docker's market success has largely focused on the needs of businesses
in deploying web applications and the potential for a lightweight
alternative to full virtualization, these features have potentially
important implications for systems research in the area of scientific
reproducibility. In this paper, I seek to set these issues in the context
of reproducibility throughout the various domain sciences where 
computation plays an ever-increasing role, but where researchers are
largely unaware of the concerns or technologies involved in making these
computations more reproducible, extensible, and portable to other researchers. 
In so doing, I highlight elements of the Docker platform that should be
of interest to both computer systems research and domain scientists.

@@

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing
to do with the technological approaches discussed here, but stems rather from a
reluctance to publish the code used to generating the results in the
first place [@Barnes_2010]. Despite extensive evidence to the contrary
[e.g. @Ince_2012], many researchers and journals continue to assume that
summary descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis.  Until such code is available in the first
place, one cannot even begin to encounter the problems that the approaches
discussed here set out to solve. As a result, few domain researchers
may be fully aware of the challenges involved in effectively re-using
published code.

@@

Studies focusing on code that has been made available with scientific
publications regularly find the same common issues that pose substantial
barriers to reproducing the original results or building on that code
[e.g. @Lapp_2014; @Garijo_2013; @Gilbert_2012; @Collberg_2014;
@Krishnamurthi_2014], which I attempt to summarize as follows.

@@

Two dominant paradigms have emerged to address these issues so far:
workflow software [e.g. @Altintas_2004; @Hull_2006]  and
cloud computing on virtual machines [e.g. @Dudley_2010; @Howe_2012]. 
Workflow software provides very elegant technical solutions to the
challenges of communication between diverse software tools, capturing
provenance in graphically driven interfaces, and handling issues from of
versioning dependencies to data access.  Workflow solutions are often 
built by well-funded collaborations between domain scientists and computer
scientists, and can be very successful in the communities within which 
they receive substantial adoption. Nonetheless, most workflow systems 
struggle with relatively low total adoption overall [@Gil_2007; @Dudley_2010].

@@

<!-- Link on how to install docker -->
<!-- Examples here are not meant as a tutorial to docker, but 
rather to illustrate concepts a reader should be able to follow whether 
or not they have any previous knowledge of docker and its commands -->

@@

With a syntax that is simpler than other provisioning tools (e.g. Chef,
Puppet, Ansible) or Continuous Integration (CI) platforms (e.g. Travis CI,
Shippable CI); users need little more than a basic familiarity with shell
scripts and a Linux distribution software environment (e.g. Debian-based
`apt-get`) to get started writing Dockerfiles.

@@

While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments [e.g. @Dudley_2010], many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.  Working locally allows a researcher to
rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces are rapidly
filling this gap).

Docker can be easily installed on most major platforms (see
[https://docs.docker.com/installation](https://docs.docker.com/installation);
On systems not already based on the Linux Kernel, such as Mac or Windows,
this is accomplished through the use of a small VirtualBox-based VM
running on the host OS called `boot2docker`) and run locally.  Docker
allows a user to link any directory (such as the working directory
for a particular script or project) to the running Docker container.
This allows a user to rely on the familiar tools of the host OS for
roles such as text editing, file browsing, or version control, while
still allowing code execution to occur inside the controlled development
environment of the container.

@@

Add the port `:8787` to the end of this address and paste
it into your browser address bar, which should open to the
RStudio welcome screen.  A user can now login with user/password
`rstudio/rstudio`, run R scripts, install packages, use git,
and so forth. User login and other configurations can be
customized using environmental variables; see details at
[https://github.com/ropensci/docker](https://github.com/ropensci/docker)

@@

One can  share a public copy of the image just created by using the
`docker push` command, followed by the name of the image using the command:

@@

If a Dockerfile is made available on a public code repository such as
[Github](https://github.com) or [Bitbucket](https://bitbucket.org),
the Hub can automatically build the image whenever a change is made to
the Dockerfile, making the `push` command unnecessary. A user can update
their local image using the `docker pull <imagename>`, which downloads
any changes that have since been made to the copy of the image on the Hub.

@@

```bash
docker build -t username/r-recommended .
```

@@

In addition to version managing the Dockerfile, the images themselves are
versioned using a git-like hash system (e.g. see `docker commit`, `docker
push`/`docker pull`, `docker history`, `docker diff`).  Docker images and
containers have dedicated metadata specifying the date, author, parent
image, and other details (see `docker inspect`). One can roll back an
image through the layers of history of its construction, then build off
an earlier layer, or roll back changes made interactively in a
container. For instance, here I inspect recent changes made to the
`ubuntu:latest` image:

@@

One can identify an earlier version, and roll back to that version 
just by adjusting the docker tag to match the hash of that version. 
For instance:

@@

This same feature also means that docker can perform incremental
uploads and downloads that send only the differences between images,
(just like `git push` or `git pull` for git repositories), rather than
transfer the full image each time.

@@

- _Use docker containers during development_. A key feature of the Docker
approach is the ability to mimic as closely as possible the current
workflow and development practices of the user.  Code executing inside
a container on a local machine can appear identical to code running
natively, but with the added benefit that one can simply recreate or
snapshot and share the entire computational environment with a few simple
commands.  This works best if researchers set up their computational
environment in a container from the outset of the project.

@@

- _Use and provide appropriate base images_. Though Docker supports
modular design, it remains up to the researchers to take advantage of
it. An appropriate workflow might involve one Dockerfile that includes
all the software dependencies a researcher usually uses in the course of
their development, which can then be extended by separate docker images
for particular projects.  Re-using existing images reduces the effort
required to set up an environment, contributes to the standardization
of computational environments within a field, and best leverages the
ability of Docker's distribution system to download only differences.

@@

- _Archive tarball snapshots_. Despite similar semantics to git, Docker's
versioning system works rather differently than version management
of code.  Docker can roll back AUFS layers that have been added to
an image, but not revert to the earlier state of a particular layer.
In consequence, to preserve a bitwise identical snapshot of a container
used to generate a given set of results, it is necessary to archive the
image tarball itself -- one can not simply rely on the docker history
to recover an earlier state.

@@

- Docker does not provide complete virtualization but relies on the 
Linux kernel provided by the host.  Systems research can provide insight
on what limitations to reproducibility this introduces [e.g.  @Harji_2013].

@@

It is possible, and indeed often desirable, to tackle the problem of
dependencies at a much higher level than that of the operating system
itself [e.g. @Ooms_2014].  For instance, it is often sufficient to
address these concerns at the level of a particular scripting language
rather than the level of the operating system. Most major scripting
languages have dedicated tools for handling dependencies and versioning
(e.g. Ruby gems, R packages, Python eggs, Perl modules, JavaScript node
package manager); with some languages having the fortune or misfortune to
have multiple systems. Though systems differ somewhat in their strategies
and their effectiveness, all are generally vulnerable to changes in their
lower-level dependencies (such as compilers, C libraries, and so forth)
that may differ between platforms.  More pointedly, researchers whose
work is not easily isolated to a single such environment but combines
tools from different systems (including combining those tools used for
formatting & visual display, such as LaTeX or pandoc, with scripting or
analysis tools such as R).

<!-- Dynamic documents --> 
Dynamic documents are another example of an approach that to reproducible
research that can be facilitated through the use of Docker containers. 
Dynamic Documents [e.g. @Leisch_2002; @Peng_2011; @Xie_2013] seek to embed
the code required to re-generate the results within the manuscript
itself. This process serves as a way to distribute computational
code with publications, to ensure that methodological descriptions,
parameter values, figures, and so forth remain synchronized with the code,
and to provide a way for other researchers to both tweak the code and
'execute' the paper to reproduce the results.  While these features
are very desirable from the perspective of reproducible research,
they inevitably increase the number of software dependencies involved,
including those tools that support combining code with text [e.g. @Xie_2013]
as well as tools merely for the handling the formating and layout
(e.g. LaTeX, pandoc).  Consequently, such a paper can fail to execute
successfully for reasons having nothing to do with the computations
involved in the research, but rather due to software responsible for
implementing these dynamic features.  This problem is largely resolved by
providing a consistent development environment, making the approaches
discussed here a natural complement to that of dynamic documents.
A similar argument can be made regarding the use of workflow systems,
which can introduce additional software dependencies even as they seek
to make it easier to work across them.

@@

I noted at the outset that cultural expectations responsible for a
lack of code sharing practices in many fields are a far more extensive
primary barrier to reproducibility than the technical barriers
discussed here. Nevertheless, it may be worth considering how solutions
to these technical barriers can influence the cultural landscape as well.
Many researchers may be reluctant to publish code today because they
fear a it will be primarily a one-way street: more technical savvy
researchers then themselves can benefit from their hard work, while
they may not benefit from the work produced by others.  Lowering the
technical barriers to reuse provides immediate practical benefits
that make this exchanged into a more balanced, two-way street. Another
concern is that the difficulty imposed in preparing code to be shared,
such as providing even semi-adequate documentation or support for other
users to be able to install and run it in the first place is too high
[@Stodden_2010]. Thus, lowering the barriers these barrier to re-use through
the appropriate infrastructure may also reduce certain cultural
barriers to sharing.

@@

Researchers familiar with LXC may be tempted to view the Docker
system as little more than a creative application and marketing of an
existing technology. However, the Docker approach adds several features
crucial for its use in reproducible research (adapted from [Docker
FAQ](https://docs.docker.com/faq/#what-does-docker-add-to-just-plain-lxc)):

>>>>

Keywords:
  - Docker
  - virtualization
  - reproducibility

@@

Reproducible research has received an increasing level of attention
throughout the scientific community [_e.g._ @Peng_2011; @Nature_2012]
and the public at large [_e.g._ @The_Economist_2013]. All steps of the
scientific process, from data collection and processing, to analyses,
visualizations and conclusions depend ever more on computation and
algorithms, _computational reproducibility_ has received particular
attention [@Merali_2010].  Though in principle this algorithmic dependence
should make such research easier to reproduce -- computer codes being
both more portable and potentially more precise to exchange and run
than experimental methods -- in practice this has led to an ever larger
and more complex black box that stands between what was actually done
and what is described in the literature.  Crucial scientific processes
such as replicating the results, extending the approach or testing the
conclusions in other contexts, or even merely installing the software
used by the original researchers can become immensely time-consuming if
not impossible.

@@

Systems research has long concerned itself with the issues of
computational reproducibility and the technologies that can facilitate
those objectives [_e.g._ @Eide_2010, @Harji_2013]. Docker is a new but already
very popular open source tool that combines many of these approaches
in a user friendly implementation, including: (1) performing Linux
container (LXC) based operating system (OS) level virtualization, (2)
portable deployment of containers across platforms, (3) component reuse,
(4) sharing, (5) archiving, and (6) versioning of container images. While
Docker's market success has largely focused on the needs of businesses
in deploying web applications and the potential for a lightweight
alternative to full virtualization, these features have potentially
important implications for systems research in the area of scientific
reproducibility. In this paper, I seek to set these issues in the context
of reproducibility throughout the various domain sciences where 
computation plays an ever-increasing role, but where researchers are
largely unaware of the concerns or technologies involved in making these
computations more reproducible, extensible, and portable to other researchers. 
In so doing, I highlight elements of the Docker platform that should be
of interest to both computer systems research and domain scientists.

@@

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing
to do with the technological approaches discussed here, but stems rather from a
reluctance to publish the code used to generating the results in the
first place [@Barnes_2010]. Despite extensive evidence to the contrary
[_e.g._ @Ince_2012], many researchers and journals continue to assume that
summary descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis.  Until such code is available in the first
place, one cannot even begin to encounter the problems that the approaches
discussed here set out to solve. As a result, few domain researchers
may be fully aware of the challenges involved in effectively re-using
published code.

@@

Studies focusing on code that has been made available with scientific
publications regularly find the same common issues that pose substantial
barriers to reproducing the original results or building on that code
[_e.g._ @Lapp_2014; @Garijo_2013; @Gilbert_2012; @Collberg_2014;
@Krishnamurthi_2014], which I attempt to summarize as follows.

@@

Two dominant paradigms have emerged to address these issues so far:
workflow software [_e.g._ @Altintas_2004; @Hull_2006]  and
cloud computing on virtual machines [_e.g._ @Dudley_2010; @Howe_2012]. 
Workflow software provides very elegant technical solutions to the
challenges of communication between diverse software tools, capturing
provenance in graphically driven interfaces, and handling issues from of
versioning dependencies to data access.  Workflow solutions are often 
built by well-funded collaborations between domain scientists and computer
scientists, and can be very successful in the communities within which 
they receive substantial adoption. Nonetheless, most workflow systems 
struggle with relatively low total adoption overall [@Gil_2007; @Dudley_2010].

@@

<!-- Link on how to install Docker -->
<!-- Examples here are not meant as a tutorial to Docker, but 
rather to illustrate concepts a reader should be able to follow whether 
or not they have any previous knowledge of Docker and its commands -->

@@

With a syntax that is simpler than other provisioning tools (_e.g._ Chef,
Puppet, Ansible) or Continuous Integration (CI) platforms (_e.g._ Travis CI,
Shippable CI); users need little more than a basic familiarity with shell
scripts and a Linux distribution software environment (_e.g._ Debian-based
`apt-get`) to get started writing Dockerfiles.

@@

While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments [_e.g._ @Dudley_2010], many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.  Working locally allows a researcher to
rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces are rapidly
filling this gap).

Docker can be easily installed on most major platforms (see
[https://docs.Docker.com/installation](https://docs.docker.com/installation);
On systems not already based on the Linux Kernel, such as Mac or Windows,
this is accomplished through the use of a small VirtualBox-based VM
running on the host OS called `boot2docker`) and run locally.  Docker
allows a user to link any directory (such as the working directory
for a particular script or project) to the running Docker container.
This allows a user to rely on the familiar tools of the host OS for
roles such as text editing, file browsing, or version control, while
still allowing code execution to occur inside the controlled development
environment of the container.

@@

Add the port `:8787` to the end of this address and paste
it into your browser address bar, which should open to the
RStudio welcome screen.  A user can now login with user/password
`rstudio/rstudio`, run R scripts, install packages, use git,
and so forth. User login and other configurations can be
customized using environmental variables; see details at
[https://github.com/ropensci/Docker](https://github.com/ropensci/docker)

@@

One can  share a public copy of the image just created by using the
`Docker push` command, followed by the name of the image using the command:

@@

If a Dockerfile is made available on a public code repository such as
[Github](https://github.com) or [Bitbucket](https://bitbucket.org),
the Hub can automatically build the image whenever a change is made to
the Dockerfile, making the `push` command unnecessary. A user can update
their local image using the `Docker pull <imagename>`, which downloads
any changes that have since been made to the copy of the image on the Hub.

@@

```bash
Docker build -t username/r-recommended .
```

@@

In addition to version managing the Dockerfile, the images themselves are
versioned using a git-like hash system (_e.g._ see `Docker commit`, `docker
push`/`Docker pull`, `docker history`, `docker diff`).  Docker images and
containers have dedicated metadata specifying the date, author, parent
image, and other details (see `Docker inspect`). One can roll back an
image through the layers of history of its construction, then build off
an earlier layer, or roll back changes made interactively in a
container. For instance, here I inspect recent changes made to the
`ubuntu:latest` image:

@@

One can identify an earlier version, and roll back to that version 
just by adjusting the Docker tag to match the hash of that version. 
For instance:

@@

This same feature also means that Docker can perform incremental
uploads and downloads that send only the differences between images,
(just like `git push` or `git pull` for git repositories), rather than
transfer the full image each time.

@@

- _Use Docker containers during development_. A key feature of the Docker
approach is the ability to mimic as closely as possible the current
workflow and development practices of the user.  Code executing inside
a container on a local machine can appear identical to code running
natively, but with the added benefit that one can simply recreate or
snapshot and share the entire computational environment with a few simple
commands.  This works best if researchers set up their computational
environment in a container from the outset of the project.

@@

- _Use and provide appropriate base images_. Though Docker supports
modular design, it remains up to the researchers to take advantage of
it. An appropriate workflow might involve one Dockerfile that includes
all the software dependencies a researcher usually uses in the course of
their development, which can then be extended by separate Docker images
for particular projects.  Re-using existing images reduces the effort
required to set up an environment, contributes to the standardization
of computational environments within a field, and best leverages the
ability of Docker's distribution system to download only differences.

@@

- _Archive tarball snapshots_. Despite similar semantics to git, Docker's
versioning system works rather differently than version management
of code.  Docker can roll back AUFS layers that have been added to
an image, but not revert to the earlier state of a particular layer.
In consequence, to preserve a bitwise identical snapshot of a container
used to generate a given set of results, it is necessary to archive the
image tarball itself -- one can not simply rely on the Docker history
to recover an earlier state.

@@

- Docker does not provide complete virtualization but relies on the 
Linux kernel provided by the host.  Systems research can provide insight
on what limitations to reproducibility this introduces [_e.g._  @Harji_2013].

@@

Using Docker containers to distribute reproducible research should be
seen as an approach that is synergistic with, rather than an alternative
to, other technical tools for ensuring computational reproducibility.
Existing tools for managing dependencies for a particular language [R
packages, Ruby gems, etc, see @Ooms_2014] can easily be employed within a
Docker-based approach, allowing the operating-systems level virtualization
to sidestep potential issues such as external library dependencies or
conflicts with existing user libraries.  Other approaches that facilitate
reproducible research also introduce additional software dependencies
and possible points of failure [@FitzJohn_2014]. One example includes
dynamic documents [_e.g._ @Leisch_2002; @Peng_2011; @Xie_2013] which embed
the code required to re-generate the results within the manuscript.  As a
result, it is necessary to package the appropriate typesetting libaries
(_e.g._ \LaTeX) along with the code libaries such that the document executes
successfully for different researchers and platforms.

@@

I noted at the outset that cultural expectations responsible for a
lack of code sharing practices in many fields are a far more extensive
primary barrier to reproducibility than the technical barriers
discussed here. Nevertheless, it may be worth considering how solutions
to these technical barriers can influence the cultural landscape as well.
Many researchers may be reluctant to publish code today because they
fear a it will be primarily a one-way street: more technical savvy
researchers then themselves can benefit from their hard work, while
they may not benefit from the work produced by others.  Lowering the
technical barriers to reuse provides immediate practical benefits
that make this exchanged into a more balanced, two-way street. Another
concern is that the difficulty imposed in preparing code to be shared,
such as providing even semi-adequate documentation or support for other
users to be able to install and run it in the first place is too high
[@Stodden_2010]. Thus, lowering these barrier to re-use through
the appropriate infrastructure may also reduce certain cultural
barriers to sharing.

@@

Researchers familiar with LXC may be tempted to view the Docker
system as little more than a creative application and marketing of an
existing technology. However, the Docker approach adds several features
crucial for its use in reproducible research (adapted from [Docker
FAQ](https://docs.Docker.com/faq/#what-does-docker-add-to-just-plain-lxc)):
